---
title: "Using R for Data Science"
author: "Bharat Khatri"
date: "Thursday 11 December 2014"
output: html_document
---

Data Types in R
---------------
1. **Numerics** 
```{r} 
my_numeric <- 42.5
```
2. **Logical** 
```{r}
my_logical <- TRUE
```
3. **Characters** 
```{r} 
my_character <- "some text"
```

Note : R is *case sensitive*

You can check the data type of a variable by using the `class()` function.

```{r}
class(my_numeric)
class(my_logical)
class(my_character)
```


Vectors
-------
R uses the combine function `c()` to instantiate vectors.

Note: Vectors in R are indexed from `1` and not `0`.

```{r}
numeric_vector <- c(42, 42.5, -1)
numeric_vector

character_vector <- c("meaning", "of", "life")
character_vector

boolean_vector <- c(FALSE, FALSE, TRUE)
boolean_vector
```

In order to augment vectors with some context, R allows you to give a *name* to each element of a vector, using the `names()` function. Vectors with names assigned to each of the elements are sometimes called *named vectors*.

```{r}
click_through_rates <- c(0.5, 0.5, 0.4, 0.2, 0.1, 0.1, 0.1)
names(click_through_rates) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
```

We use the bracket notation `[]` to access elements of a vector.

```{r}
click_through_rate_tue <- click_through_rates["Tue"]
click_through_rate_tue
```

To access multiple elements of a vector together, you need to use a vector inside the bracket notation.

```{r}
click_through_rates_weekday <- click_through_rates[c("Mon", "Tue", "Wed", "Thu", "Fri")]
click_through_rates_weekday
```

There's another way you can select specific elements from a vector.

```{r}
click_through_rates_weekday <- click_through_rates[c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE)]
click_through_rates_weekday
```

R supplements vector data type with `aggregate` functions.

```{r}
daily_expenses <- c(100, 80, 120, 100, 500, 1000, 500)
names(daily_expenses) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

total_expense <- sum(daily_expenses)
total_expense

mean_expense <- mean(daily_expenses)
mean_expense
```


Note : There is a shorthand for generating a vector of contiguous integers.
```{r}
num_campus_hires <- c(0, 1, 2, 3, 4)
num_campus_hires

num_campus_hires <- 0:4
num_campus_hires
```

Map operations on vectors
-------------------------
```{r}
rpms <- c(1.2, 2.3, -1, 15.1, -1, 5.7)
names(rpms) <- c("thekrazycouponlday.com", "obamacarefacts.com", "amazon.com", "goodhousekeeping.com", "wikipedia.org", "cheapflights.com")

rpm_predictor_failures <- rpms < 0
rpm_predictor_failures

crap_urls <- names(rpms)[rpm_predictor_failures]
crap_urls
```


Matrices
--------
`Definition` : In R, a matrix is a two-dimensional collection of elements of the same data type (*numeric*, *character*, or *logical*)

Matrices can be constructed using the function `matrix()`

```{r}
# Fill c(1,2,3,4,5,6,7,8,9) row wise in a 3x3 square matrix
row_wise_matrix <- matrix(1:9, byrow=TRUE, nrow=3)
row_wise_matrix

# Fill c(1,2,3,4,5,6,7,8,9) column wise in a 3x3 square matrix
col_wise_matrix <- matrix(1:9, byrow=FALSE, nrow=3)
col_wise_matrix
```


Naming matrices
---------------
As with vectors, R allows you to name the *rows* and *columns* of a matrix using `rownames()` and `colnames()` respectively. Naming rows and columns helps both in associating a context with the data, as well as accessing data using better understood labels. 

```{r}
weekend_ctrs <- matrix(c(7.1, 6.9, 0.8, 0.7, 1.8, 1.6), byrow=TRUE, nrow=3)
rownames(weekend_ctrs) <- c("goodhousekeeping.com", "thekrazycouponlady.com", "cheapflights.com")
colnames(weekend_ctrs) <- c("Sat", "Sun")
weekend_ctrs
```


Factors
---------------
The term factor refers to a statistical data type used to store categorical variables. It stores the information about the variable (i.e the values it can take).
  Suppose we have a vector of observations that belong to a limited number of categories. For example, gender_vector below contains the sex of 5 different individuals

```{r}
gender_vector <- c("Male","Female","Female","Male","Male")
```

You can create factor by using the function factor() and see what it variable it stores.

```{r}
factor_gender_vector <- factor(gender_vector)
```

There are two types of categorical variables. A nominal categorical variable and an ordinal categorical variable. To impose order on categories you can use factor function in the below way

```{r}
temperature_vector <- c("Hig","Low","Hig","Low","Med")
factor_temperature_vector <- factor(temperature_vector, order=TRUE, levels=c("Low","Medium","High"))

```

You can manually rename the levels for your convinience

```{r}
levels(factor_temperature_vector) <- c("Low","Medium","High")
```

You can use summary() function to see the number of elements of each category

```{r}
summary(factor_temperature_vector)
```

Now you can access the underlying vector of observations through the factor

```{r}
factor_temperature_vector[1]
factor_temperature_vector[2]  
```

When the factor is ordinal you can compare two observations for the order

```{r}
factor_temperature_vector[1] < factor_temperature_vector[2]  
```


One More Example

```{r}
speed_vector <- c("Fast","Slow","Slow","Fast","Ultra-fast")
factor_speed_vector <- factor(speed_vector, ordered=TRUE, levels=c("Slow","Fast","Ultra-fast"))
compare_them <- factor_speed_vector[2] > factor_speed_vector[5]
compare_them
```


Data Frames
----------------
Matrix restricts all the elements to be of same type but that's not the case all times. For this dataframes are used. Below is an example dataset stored in R by default

```{r}
mtcars
```

You can have a peek at data using functions head(),tail()

```{r}
head(mtcars)
tail(mtcars)
```

To get a initial insight into data you can use str() function. str() shows you the structure of the data frame. str() gives the information about no of entries, no of variables, type of each variable and first few entries for each variable.

```{r}
str(mtcars)
```

Data Frames Creation :
--------------

Lets say you have column vectors for different variables. Each vector has it's own data type. You can create the data frame as below.

```{r}
planets     <- c("Mercury","Venus","Earth","Mars","Jupiter","Saturn","Uranus","Neptune");
type        <- c("Terrestrial planet","Terrestrial planet","Terrestrial planet","Terrestrial planet","Gas giant","Gas giant","Gas giant","Gas giant")
diameter    <- c(0.382,0.949,1,0.532,11.209,9.449,4.007,3.883); 
rotation    <- c(58.64,-243.02,1,1.03,0.41,0.43,-0.72,0.67);
rings       <- c(FALSE,FALSE,FALSE,FALSE,TRUE,TRUE,TRUE,TRUE);

# Create the data frame:
planets_df  <- data.frame(planets,type,diameter,rotation,rings)
```

You can see the overview of planets_df using str() function
```{r}
str(planets_df)
```

Similar to vectors and matrices, you select elements from a data frame with the help of square brackets [ ]. By using a comma, you can indicate what to select from the rows and the columns respectively. 

```{r}
planets_df[1,2]
planets_df[1:3,2:4]
planets_df[1,]
planets_df[1:3]
planets_df[3:1,]
planets_df[3:1,1]
planets_df[3:1,"type"]
furthest_planets_diameter <- planets_df[3:8,"diameter"]
```

To extract a column from data frame you can use [,column_num] or [,column_name ] but there is another way using $ symbol


```{r}
planets_df$diameter
```

You can use subset() function to extract a subset of variables using conditions

```{r}
subset(planets_df,subset=(planets_df$rings==TRUE))
```
Sorting is one of the important function used in data analysis. In R order() is used for sorting. order() when applied on an array will give position of each element when they are sorted



```{r}
a <- c(1,0.4,3,2.1,2.2)
order(a)
a[order(a)]
```

