---
title: "Using R for Data Science"
author: "Bharat Khatri"
date: "Thursday 11 December 2014"
output: html_document
---

Data Types in R
---------------
1. **Numerics** 
```{r} 
my_numeric <- 42.5
```
2. **Logical** 
```{r}
my_logical <- TRUE
```
3. **Characters** 
```{r} 
my_character <- "some text"
```

Note : R is *case sensitive*

You can check the data type of a variable by using the `class()` function.

```{r}
class(my_numeric)
class(my_logical)
class(my_character)
```


Vectors
-------
R uses the combine function `c()` to instantiate vectors.

Note: Vectors in R are indexed from `1` and not `0`.

```{r}
numeric_vector <- c(42, 42.5, -1)
numeric_vector

character_vector <- c("meaning", "of", "life")
character_vector

boolean_vector <- c(FALSE, FALSE, TRUE)
boolean_vector
```

In order to augment vectors with some context, R allows you to give a *name* to each element of a vector, using the `names()` function. Vectors with names assigned to each of the elements are sometimes called *named vectors*.

```{r}
click_through_rates <- c(0.5, 0.5, 0.4, 0.2, 0.1, 0.1, 0.1)
names(click_through_rates) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
```

We use the bracket notation `[]` to access elements of a vector.

```{r}
click_through_rate_tue <- click_through_rates["Tue"]
click_through_rate_tue
```

To access multiple elements of a vector together, you need to use a vector inside the bracket notation.

```{r}
click_through_rates_weekday <- click_through_rates[c("Mon", "Tue", "Wed", "Thu", "Fri")]
click_through_rates_weekday
```

There's another way you can select specific elements from a vector.

```{r}
click_through_rates_weekday <- click_through_rates[c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE)]
click_through_rates_weekday
```

R supplements vector data type with `aggregate` functions.

```{r}
daily_expenses <- c(100, 80, 120, 100, 500, 1000, 500)
names(daily_expenses) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

total_expense <- sum(daily_expenses)
total_expense

mean_expense <- mean(daily_expenses)
mean_expense
```


Note : There is a shorthand for generating a vector of contiguous integers.
```{r}
num_campus_hires <- c(0, 1, 2, 3, 4)
num_campus_hires

num_campus_hires <- 0:4
num_campus_hires
```

Map operations on vectors
-------------------------
```{r}
rpms <- c(1.2, 2.3, -1, 15.1, -1, 5.7)
names(rpms) <- c("thekrazycouponlday.com", "obamacarefacts.com", "amazon.com", "goodhousekeeping.com", "wikipedia.org", "cheapflights.com")

rpm_predictor_failures <- rpms < 0
rpm_predictor_failures

crap_urls <- names(rpms)[rpm_predictor_failures]
crap_urls
```
