---
title: "Using R for Data Science"
author: "Bharat Khatri"
date: "Thursday 11 December 2014"
output: html_document
---

Data Types in R
---------------
1. **Numerics** 
```{r} 
my_numeric <- 42.5
```
2. **Logical** 
```{r}
my_logical <- TRUE
```
3. **Characters** 
```{r} 
my_character <- "some text"
```

Note : R is *case sensitive*

You can check the data type of a variable by using the `class()` function.

```{r}
class(my_numeric)
class(my_logical)
class(my_character)
```


Vectors
-------
R uses the combine function `c()` to instantiate vectors.

Note: Vectors in R are indexed from `1` and not `0`.

```{r}
numeric_vector <- c(42, 42.5, -1)
numeric_vector

character_vector <- c("meaning", "of", "life")
character_vector

boolean_vector <- c(FALSE, FALSE, TRUE)
boolean_vector
```

In order to augment vectors with some context, R allows you to give a *name* to each element of a vector, using the `names()` function. Vectors with names assigned to each of the elements are sometimes called *named vectors*.

```{r}
click_through_rates <- c(0.5, 0.5, 0.4, 0.2, 0.1, 0.1, 0.1)
names(click_through_rates) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
```

We use the bracket notation `[]` to access elements of a vector.

```{r}
click_through_rate_tue <- click_through_rates["Tue"]
click_through_rate_tue
```

To access multiple elements of a vector together, you need to use a vector inside the bracket notation.

```{r}
click_through_rates_weekday <- click_through_rates[c("Mon", "Tue", "Wed", "Thu", "Fri")]
click_through_rates_weekday
```

There's another way you can select specific elements from a vector.

```{r}
click_through_rates_weekday <- click_through_rates[c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE)]
click_through_rates_weekday
```

R supplements vector data type with `aggregate` functions.

```{r}
daily_expenses <- c(100, 80, 120, 100, 500, 1000, 500)
names(daily_expenses) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

total_expense <- sum(daily_expenses)
total_expense

mean_expense <- mean(daily_expenses)
mean_expense
```


Note : There is a shorthand for generating a vector of contiguous integers.
```{r}
num_campus_hires <- c(0, 1, 2, 3, 4)
num_campus_hires

num_campus_hires <- 0:4
num_campus_hires
```

Map operations on vectors
-------------------------
```{r}
rpms <- c(1.2, 2.3, -1, 15.1, -1, 5.7)
names(rpms) <- c("thekrazycouponlday.com", "obamacarefacts.com", "amazon.com", "goodhousekeeping.com", "wikipedia.org", "cheapflights.com")

rpm_predictor_failures <- rpms < 0
rpm_predictor_failures

crap_urls <- names(rpms)[rpm_predictor_failures]
crap_urls
```


Matrices
--------
`Definition` : In R, a matrix is a two-dimensional collection of elements of the same data type (*numeric*, *character*, or *logical*)

Matrices can be constructed using the function `matrix()`

```{r}
# Fill c(1,2,3,4,5,6,7,8,9) row wise in a 3x3 square matrix
row_wise_matrix <- matrix(1:9, byrow=TRUE, nrow=3)
row_wise_matrix

# Fill c(1,2,3,4,5,6,7,8,9) column wise in a 3x3 square matrix
col_wise_matrix <- matrix(1:9, byrow=FALSE, nrow=3)
col_wise_matrix
```


Naming matrices
---------------
As with vectors, R allows you to name the *rows* and *columns* of a matrix using `rownames()` and `colnames()` respectively. Naming rows and columns helps both in associating a context with the data, as well as accessing data using better understood labels. 

```{r}
weekend_ctrs <- matrix(c(7.1, 6.9, 0.8, 0.7, 1.8, 1.6), byrow=TRUE, nrow=3)
rownames(weekend_ctrs) <- c("goodhousekeeping.com", "thekrazycouponlady.com", "cheapflights.com")
colnames(weekend_ctrs) <- c("Sat", "Sun")
weekend_ctrs
```

You can add additional rows and columns to an existing matrix using `rbind()` and `cbind()` respectively.

```{r}

# Binding matrices by column
weekday_ctrs <- matrix(c(14.1, 15.3, 14, 15.1, 12, 1.1, 1.4, 1.5, 1.1, 1.1, 2.1, 2.4, 2.2, 2.3, 2.1), byrow=TRUE, nrow=3)
rownames(weekday_ctrs) <- c("goodhousekeeping.com", "thekrazycouponlady.com", "cheapflights.com")
colnames(weekday_ctrs) <- c("Mon", "Tue", "Wed", "Thu", "Fri")
weekday_ctrs

week_ctrs <- cbind(weekday_ctrs, weekend_ctrs)
week_ctrs

# Binding matrices by row
new_weekend_ctrs <- matrix(c(3.1, 2.7, 6.2, 6.1), byrow=TRUE, nrow=2)
rownames(new_weekend_ctrs) <- c("obamacarefacts.com", "finance.yahoo.com")
colnames(new_weekend_ctrs) <- c("Sat", "Sun")
new_weekend_ctrs

all_weekend_ctrs <- rbind(weekend_ctrs, new_weekend_ctrs)
all_weekend_ctrs
```

R also supplements matrices with `aggregate` functions.

```{r}
all_expenses = matrix(c(100, 60, 20, 100, 300, 500, 0, 100, 120, 120, 100, 100, 0, 600, 50, 60, 50, 20, 30, 40, 100, 0, 0, 0, 0, 0, 1800, 0, 300, 30, 20, 50, 80, 100, 100), byrow=TRUE, nrow=5)
rownames(all_expenses) <- c("Food", "Travel", "Communication", "Clothing", "Miscellaneous")
colnames(all_expenses) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
all_expenses

category_wise_expenses <- rowSums(all_expenses)
category_wise_expenses

day_wise_expenses <- colSums(all_expenses)
day_wise_expenses
```

R uses bracket notation `[]` to access elements from a matrix.

```{r}
sun_miscellaneous_expenses <- all_expenses["Miscellaneous", "Sun"]
food_expenses <- all_expenses["Food",]
mon_expenses <- all_expenses[, "Mon"]
```


Factors
-------
Factors are used to store *categorical* variables. It is important that R knows whether it is dealing with a continous or a categorical variable, as the statistical models you will develop in the future treat both types differently. 

Factors are created using `factor()`

```{r}
visitor_countries <- c("Tier 1", "Tier 2", "Tier 1", "Tier 3", "Tier 1")
factor_visitor_countries <- factor(visitor_countries)
factor_visitor_countries
```

There are two types of categorical variables :

1. **Nominal** categorical variable
2. **Ordinal** categorical variable

Nominal variable

Nominal categorical variables do not have an implied order. 

```{r}
adult_filter_values <- c("ADULT", "NON-ADULT", "NON-ADULT", "NON-ADULT")
nominal_variables <- factor(adult_filter_values)
nominal_variables 
```

Ordinal variable

Ordinal categorical variables have a natural order.

```{r}
in_times <- c("EARLY", "REGULAR", "REGULAR", "LATE", "LATE", "LATE", "LATE")
ordinal_variables <- factor(in_times, order=TRUE, levels=c("LATE", "REGULAR", "EARLY"))
ordinal_variables
```

R allows you to *rename* the levels of factors.

```{r}
visitor_country_tiers <- c("1", "1", "2", "1", "3")
factor_visitor_country_tiers <- factor(visitor_country_tiers)
factor_visitor_country_tiers

levels(factor_visitor_country_tiers)

levels(factor_visitor_country_tiers) <- c("America", "Asia", "Africa")
levels(factor_visitor_country_tiers)

factor_visitor_country_tiers
```


Summarizing a factor
--------------------
`summary()` is one of the most useful functions of R, can can be used to get a quick overview of any R variable. Let's see what `summary()` provides us for factors.

```{r}
in_times <- c("EARLY", "REGULAR", "REGULAR", "LATE", "LATE", "LATE", "LATE")
factor_in_times <- factor(in_times, order=TRUE, levels=c("LATE", "REGULAR", "EARLY"))
summary(factor_in_times)
```

Comparision operations on ordinal factors
-----------------------------------------
```{r}
in_times <- c("EARLY", "REGULAR", "REGULAR", "LATE", "LATE", "LATE", "LATE")
names(in_times) <- c("Vinay", "Rajat", "Sachin", "Sonkar", "Bharat", "Rishabh", "Vinod")
in_times

factor_in_times <- factor(in_times, order=TRUE, levels=c("LATE", "REGULAR", "EARLY"))
factor_in_times

factor_in_times["Sachin"] > factor_in_times["Sonkar"]
# Comparing factor elements
```

Data Frames
-----------
A data frame has the attributes of a dataset as columns and the observations as rows. A data frame is used as the fundamental data structure by most of R's modeling software.


`head()` and `tail()` allow you to inspect data frames.

```{r}
# Access 10 elements from head
head(mtcars, n=10)

# Access 15 elements from tail
tail(mtcars, n=15)
```

`str()` allows you to get a good summary of your dataset. This would often be the first thing that you do when you receive a data set.

```{r}
str(mtcars)
```

Constructing data frames
```{r}
keyword_categories <- factor(c("LEARNED_PATTERN_URL", "SERVER_SIDE_BEHAVIORAL", "AUTO_TAXONOMY_BROAD", "POPCAT_PATTERN", "NON_KBB"))                       
```


Built-in datasets in R
----------------------
1. mtcars
2. iris

ARFF (Attribute Relation File Format)
-------------------------------------
An ARFF file is an ASCII text file descrbing a list of instances sharing a set of attributes.
ARFF files are composed of two sections: 

1. **Header** (contains the name of the relation, a list of attributes and their types)
2. **Data** (contains the instances following the relation)

Attribute data types
--------------------
1. Numeric (includes *integer* and *real*)
2. Categorical (nominal or ordinal)
3. String (can contain arbitrary textual values)
4. Date

Data Science Experiments
------------------------

Procedure :
1. Data Preparation
  - Handling missing values
  - Feature selection
  - Train/Test split
2. Pre processing
  - Dimensionality Reduction
3. Model Selection
4. Model Validation
  - Cross validation
5. Model Accuracy
6. Model Results
  - Summary
    - Confusion Matrix
  

Why is cross-validation useful?
- N-fold cross validation


Measures of Model Performance
-----------------------------
1. Precision
2. Recall
3. F-measure
4. ROC Area
5. Kappa Statistic

Naive Bayes
-----------
We'll be using the e1071 R package to construct the naive bayes classifier.
```{r}
install.packages('e1071')
library('e1071')
classifier <- naiveBayes(iris[, 1:4], iris[, 5])

# Confusion Matrix
table(predict(classifier, iris[,-5]), iris[,5], dnn=list('predicted','actual'))


# Summary
classifier


classifier$tables$Petal.Length


# Plot
plot(function(x) dnorm(x, 1.462, 0.1736640), 0, 8, col="red", main="Petal length distribution for the 3 different species")
curve(dnorm(x, 4.260, 0.4699110), add=TRUE, col="blue")
curve(dnorm(x, 5.552, 0.5518947 ), add=TRUE, col = "green")
```

